\chapter{\abstractname}

Conversational agents, or chatbots,
have become a key component of modern software ecosystems,
yet their complex, non-deterministic, and heterogeneous nature
presents significant challenges for quality assurance.
Existing testing methodologies
often require source code access,
depend on manually created test scripts or pre-existing conversational corpora,
or are tied to specific development frameworks.
These limitations create a significant bottleneck
for the automated, black-box testing of deployed systems.

This thesis presents \acf{TRACER},
a methodology and open-source framework
for the automated exploration, modelling, and profiling of conversational agents.
\ac{TRACER} employs a \ac{LLM}-powered agent
to systematically interact with a deployed chatbot in a black-box manner,
automatically inferring a structured functional model of its capabilities.
This model captures the chatbot's
conversational workflows, supported parameters, and expected outputs.
From this inferred model,
\ac{TRACER} then automatically synthesises
a comprehensive set of detailed test user profiles,
designed for execution by a user simulator.
The entire methodology is supported by a full-stack web application
that provides easy access to the end-to-end workflow
from model inference with \ac{TRACER} to test execution with the SENSEI user simulator.

We validate the effectiveness of the approach through three research questions.
In a controlled, white-box environment using four open-source chatbots,
the first experiment demonstrated that \ac{TRACER} achieves high functional coverage.
In the second experiment,
the synthesised profiles successfully detected 84.6\% of injected faults in a mutation testing analysis.
In a black-box evaluation against five real-world, deployed chatbots,
the models inferred by \ac{TRACER} achieved an average precision of 96.0\% upon manual verification.
The results demonstrate that \ac{TRACER} provides a robust, practical, and effective solution
that significantly advances the state of the art in automated, black-box testing for modern conversational agents.

\vspace{2cm}

\noindent\textbf{Keywords:} Conversational Agents, Software Testing, Black-Box Testing, Automated Exploration, Test Profile Generation, Large Language Models (LLMs).
